---
title: "Ecological Genomics Final Project"
author: "Aidan Jones"
date: "2025-11-01"
output: html_document
---

## Ecological Genomics Final Project - Transposable Element Enrichment

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/users/a/j/ajones76/projects/eco_genomics_2025/final_project/mydata")
```

### Import required libraries and dependencies

```{r, message = FALSE, warning = FALSE}
# Data manipulation and plotting functions
library(tidyverse)

# Data frame construction
library(tibble)

# Clean annotation fields
library(stringr)

# Differential gene expression calculations
library(DESeq2)

# Genomic interval operations
library(GenomicRanges)

# Import sequence coordinates
library(rtracklayer)

# Sequence parsing and manipulation
library(Biostrings)

# Sequence motif scanning
library(motifmatchr)

# JASPAR interface to convert motif objects
library(TFBSTools)

# Generate clustering heatmaps
library(pheatmap)
```

## Import counts matrix and sample distibution table

```{r, warning = FALSE}
# Import the counts matrix
countsTable <- read.table("DESeq2/counts_matrix.txt", header=TRUE, row.names=1)

# Round counts to integers such that they are compatible with DESeq2
countsTableRound <- round(countsTable)

# Import the sample description table
conds <- read.delim("DESeq2/metadata.txt", 
                    header=TRUE, 
                    stringsAsFactors = TRUE, 
                    row.names=1)

# Correct column names to match between the metadata and the counts matrix
colnames(countsTableRound) <- substr(colnames(countsTableRound), 
                                     start = 1, 
                                     stop = 4)
```

### Run DESeq2 to test for differential gene expression

```{r, warning = FALSE}
# Create a DESeq object and define the experimental design
dds <- DESeqDataSetFromMatrix(countData = countsTableRound, colData=conds, 
                              design= ~ generation + line + generation:line)

# Filter out genes with too few reads (genes with less than 15 counts in more than 75% of samples)
dds <- dds[rowSums(counts(dds) >= 15) >= 18,]

# Relevel factors by defining reference for line and generation
dds$line <- relevel(dds$line, ref = "control")
dds$generation <- relevel(dds$generation, ref = "G1")

# Run the DESeq model to test for differential gene expression
dds$group <- factor(paste0(dds$line, dds$generation))
design(dds) <- ~ group
dds <- DESeq(dds)

# List the result names generated
resultsNames(dds)
```

### Set up contrast and extract DEGs

```{r, warning = FALSE}
# Set up contrast between control G1 and treatment G1
resG1_CvT <- results(dds, 
                     contrast=c("group","controlG1","treatmentG1"), 
                     alpha = 0.05)

# Clean results table
resG1_CvT <- resG1_CvT[order(resG1_CvT$padj), ]
resG1_CvT <- resG1_CvT[!is.na(resG1_CvT$padj), ]

# Extract DEGs (padj < 0.05)
degs_G1_CvT <- rownames(resG1_CvT[resG1_CvT$padj < 0.05, ])
```

### Import gene localities and TE annotations

```{r, warning = FALSE}
# Import minimap2 BED with genomic positions for every gene mapped to the reference genome from the reference transcriptome (generated from tonsa_mapping.sh)
genes_df <- read.table(
  "MiniMap2/trinity_to_genome.bed", sep="\t", header=FALSE,
  col.names=c("chr", "start", "end", "gene", "score", "strand")
)

# Keep genes that passed DESeq2 filtering
dds_genes <- rownames(dds)
genes_df <- genes_df[genes_df$gene %in% dds_genes, ]

# Convert dds gene list into GRanges object to store genes as genomic intervals
genes_gr <- GRanges(
  seqnames = genes_df$chr,
  ranges = IRanges(start = genes_df$start, end = genes_df$end),
  strand = genes_df$strand,
  gene = genes_df$gene
)

# Import RepeatMasker GFF with IDs, family associations, and genomic positions for every repetitive element masked by RepeatMasker and store as GRanges object
tes_gr <- import.gff1("RepeatMasker/acar2_round_2_trimmed.fasta.out.gff")
```

### Wilcoxon test for TE proximity

```{r, warning = FALSE}
# Compute the distance from each gene to its nearest TE annotation
dist_nearest <- distanceToNearest(genes_gr, tes_gr)

# Create a data frame of distance from TE annotation for each gene
gene_distances <- data.frame(
  gene = genes_gr$gene[queryHits(dist_nearest)],
  distance = mcols(dist_nearest)$distance
)

# Label which genes are DEGs in the data frame
gene_distances$is_DEG <- gene_distances$gene %in% degs_G1_CvT

# Run Wilcoxon test to test whether DEGs are closer to TEs than non-DEGs
wilcox.test(distance ~ is_DEG, data = gene_distances)
```

### Assign TE families

```{r, warning = FALSE}
# Assign TE annotation GRanges object as a dataframe
tes_df <- as.data.frame(tes_gr)

# Clean TE data frame to extract annotation IDs and motif names
tes_df <- tes_df %>%
  mutate(
    ID = str_extract(group, "(?<=ID=)\\d+"),
    motif_raw = str_extract(group, '(?<=Target ")[^"]+'),
    motif = str_remove(motif_raw, "^Motif:")
  ) %>%
  select(-group, -motif_raw)

# Remove TE annotations without motif names
tes_df <- tes_df[!is.na(tes_df$motif), ]

# Exclude simple repeats and low-complexity regions
tes_df <- tes_df[!grepl("^\\(", tes_df$motif), ]
tes_df <- tes_df[!grepl("^[ATCG]+-rich", tes_df$motif), ]

# Exclude structural elements
tes_df <- tes_df[!grepl("^tRNA", tes_df$motif), ]
tes_df <- tes_df[!grepl("^7S", tes_df$motif), ]
tes_df <- tes_df[!grepl("^5S", tes_df$motif), ]
tes_df <- tes_df[!grepl("^[LSUSSU]+-rRNA", tes_df$motif), ]

# Filter TE families based on minimum occurrence in the genome
min_TE_count <- 10
te_counts <- table(tes_df$motif)
keep_classes <- names(te_counts[te_counts >= min_TE_count])
tes_df <- tes_df[tes_df$motif %in% keep_classes, ]

# Add updated information to the TE annotation GRanges object
tes_gr <- GRanges(
  seqnames = tes_df$seqnames,
  ranges = IRanges(start = tes_df$start, end = tes_df$end),
  strand = tes_df$strand,
  ID = tes_df$ID,
  TE_family = tes_df$motif
)
```

### Create windows and identify overlaps with genes and TE families

```{r, warning = FALSE}
# Define promoter region size
promoter_size <- 5000

# Extend gene bodies by concatenating promoter region with strand awareness
gene_windows <- promoters(genes_gr,
                          upstream = promoter_size,
                          downstream = width(genes_gr))

# Split TE annotations by defined family
te_families <- split(tes_gr, tes_gr$TE_family)

# Build presence/absence table for each TE family near each gene
gene_TE_presence_matrix <- lapply(names(te_families), function(cl) {
  
  # Select TE annotations belonging to a single TE family
  tes_sub <- te_families[[cl]]
  
  # Find overlaps between genes and the TE family
  hits <- findOverlaps(gene_windows, tes_sub)
  
  # Create a logical vector indicating which genes have the TE family
  has_TE_family <- logical(length(genes_gr))
  if (length(hits) > 0) {
    has_TE_family[queryHits(hits)] <- TRUE
  }
  
  # Return a data frame for the TE family
  data.frame(
    gene = genes_gr$gene,
    TE_family = cl,
    has_TE = has_TE_family
  )
})

# Combine all TE families into a single data frame
gene_TE_df <- bind_rows(gene_TE_presence_matrix)

# Add DEG status
gene_TE_df <- gene_TE_df %>%
  mutate(is_DEG = gene %in% degs_G1_CvT)
```

### Run Fisher's exact tests for each TE family

```{r, warning = FALSE}
# Define function that takes the DEG vector list and the TE hit data frame
fisher_function <- function(gene_TE_df){

  # Group TE hit data frame by family
  raw_results <- gene_TE_df %>%
    group_by(TE_family) %>%
    group_modify(~{
      
      # Build contingency table
      tab <- table(.x$is_DEG, .x$has_TE)
      
      # Extract counts for each quadrant
      DEG_TE <- if(
        "TRUE" %in% rownames(tab) & 
        "TRUE" %in% colnames(tab))tab["TRUE","TRUE"] else 0
      DEG_noTE <- if(
        "TRUE" %in% rownames(tab) & 
        "FALSE" %in% colnames(tab)) tab["TRUE","FALSE"] else 0
      nonDEG_TE <- if(
        "FALSE" %in% rownames(tab) & 
        "TRUE" %in% colnames(tab)) tab["FALSE","TRUE"] else 0
      nonDEG_noTE <- if(
        "FALSE" %in% rownames(tab) & 
        "FALSE" %in% colnames(tab)) tab["FALSE","FALSE"] else 0

      # Run Fisher's exact test for complete tables 
      if(all(dim(tab) == c(2,2))) {
        ft <- fisher.test(tab)
        data.frame(
          p_value = ft$p.value,
          odds_ratio = as.numeric(ft$estimate),
          DEG_TE = DEG_TE,
          DEG_noTE = DEG_noTE,
          nonDEG_TE = nonDEG_TE,
          nonDEG_noTE = nonDEG_noTE,
          valid_test = TRUE
        )
      } else {
        data.frame(
          p_value = NA,
          odds_ratio = NA,
          DEG_TE = DEG_TE,
          DEG_noTE = DEG_noTE,
          nonDEG_TE = nonDEG_TE,
          nonDEG_noTE = nonDEG_noTE,
          valid_test = FALSE
        )
      }
    }) %>%
    ungroup()

  # Apply Benjamini–Hochberg FDR correction
  raw_results <- raw_results %>%
    mutate(p_adj = ifelse(valid_test,p.adjust(p_value, method = "fdr"), NA))
  
  return(raw_results)
}

# Run the Fisher's exact test function
fisher_results <- fisher_function(gene_TE_df = gene_TE_df)
```

### Barplot of enriched TE families near DEGs

```{r, warning = FALSE}
# Mapping TE families with broader transposon classes
te_class_map <- data.frame(
  TE_family = c("PiggyBac_A.marissinica_182",
                "hAT-N137_DR",
                "PIF-Harbinger_A.marissinica_172",
                "TE2948_SO2_FAM1039",
                "DF003450183",
                "Helitron-like-6a_Hmel",
                "DF003447903",
                "DF003448132",
                "LINE_P.maximus_88_cons"),
  TE_class = c("DNA Transposon", 
               "DNA Transposon", 
               "DNA Transposon", 
               "Unclassified", 
               "Unclassified", 
               "DNA Transposon", 
               "Unclassified",
               "Unclassified",
               "LINE")
)

# Prepare Fisher's exact test results for plotting
fisher_plot_df <- fisher_results %>%
  mutate(
    log2_or = log2(odds_ratio)) %>%
  left_join(te_class_map, by = "TE_family") %>% 
  filter(p_adj < 0.05)

# Define colors for the broad categories
te_class_colors <- c(
  "LINE" = "red",
  "DNA Transposon" = "blue",
  "Unclassified" = "grey50"
)

# Plot enriched TE families
ggplot(fisher_plot_df,
       aes(x = reorder(TE_family, log2_or),
           y = log2_or,
           fill = TE_class)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = te_class_colors, name = "TE Class") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Enriched TE Families Near DEGs (FDR < 0.05)",
    y = "log2(Odds Ratio)",
    x = "TE Family"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 10)
  )
```

### Extract TE sequences from enriched TE families

```{r, warning = FALSE}
# Identify significantly enriched TE families
enriched_families <- fisher_results %>% 
  filter(p_adj < 0.05) %>% 
  pull(TE_family) %>%
  unique()

# Load the reference genome
genome <- readDNAStringSet("RepeatMasker/acar2_round_2_trimmed.fasta")

# Split all TE annotation genomic ranges by TE family
te_by_family <- split(tes_gr, tes_gr$TE_family)

# Prepare an empty list to store TE sequences
te_sequences <- list()

# Loop through each enirhced TE family
for (fam in enriched_families) {
  
  # Retrieve GRanges intervals for the sequences of the TE family
  gr <- te_by_family[[fam]]
  
  # Extract corresponding DNA sequences from the genome
  seqs <- getSeq(genome, gr)

  # Label each DNA equence 
  names(seqs) <- paste0(
    fam, "_", 
    gr$TE_ID, "_", 
    seqnames(gr), ":", 
    start(gr), "-", end(gr))

  # Store in the TE sequences list
  te_sequences[[fam]] <- seqs
  
  # Write FASTA containing TE sequences
  writeXStringSet(seqs, 
                  filepath = paste0("MotifMatchr", 
                                    "/", 
                                    "TE_family_", 
                                    fam, 
                                    ".fasta"))
}
```

### Scan TE sequences for know invertebrate TF binding motifs

```{r, warning = FALSE}
# Load JASPAR invertebrate TF motif position frequency matrix
jaspar_file <- "MotifMatchr/JASPAR2026_CORE_insects_non-redundant_pfms_jaspar.txt"
pfm_list <- readJASPARMatrix(jaspar_file, matrixClass = "PFM")

# Extract the JASPAR IDs from each motif 
motif_ids <- sapply(pfm_list, function(x) ID(x))
names(pfm_list) <- motif_ids

# Convert position frequency matrix to position weight matrix
jaspar_pwms <- toPWM(pfm_list)

# Prepare an empty list to store motif hits
motif_hits <- list()

# Scan extracted sequences for each TE family for TF motif hits
for (fam in names(te_sequences)) {
  
  # Retrieves DNA sequences for the TE family
  seqs <- te_sequences[[fam]]
  
  # Scan each TE sequence copy against all JASPAR PWMs
  hits <- matchMotifs(
    pwm = jaspar_pwms,
    subject = seqs,
    out = "matches",
    p.cutoff = 1e-4
  )
  
  # Save binary matrix of motif matches per TE sequence
  motif_hits[[fam]] <- hits
}

# Convert motif hits matrix to a tidy summary table
motif_summary <- map_df(names(motif_hits), function(fam) {
  mat <- assay(motif_hits[[fam]], "motifMatches")
  rownames(mat) <- names(jaspar_pwms)[1:nrow(mat)]
  
  # Build table outlining the number of TE copies that contain a specific motif
  tibble(
    TE_family = fam,
    motif_id = rownames(mat),
    n_TE_copies = rowSums(mat))
  }
)

# Collapse potential motif duplicates
motif_counts <- motif_summary %>%
  group_by(TE_family, motif_id) %>%
  summarise(n_TE_copies = sum(n_TE_copies), .groups = "drop")

# Create a wide presence/absence matrix for each motif in every enriched TE family
motif_matrix <- motif_counts %>%
  mutate(present = ifelse(n_TE_copies > 0, 1, 0)) %>%
  select(TE_family, motif_id, present) %>%
  pivot_wider(names_from = motif_id, values_from = present, values_fill = 0)

# Remove motifs that do not appear in any TE family
motif_matrix_filtered <- motif_matrix %>%
  select(TE_family, which(colSums(.[,-1]) > 0) + 1)
```

### Investigate TF motif abundance in enriched TE families

```{r}
# Identify top 25 motifs across all enriched TE families based on total abundance
top_motifs <- motif_summary %>%
  filter(!is.na(motif_id)) %>%
  group_by(motif_id) %>%
  summarise(total_hits = sum(n_TE_copies), .groups = "drop") %>%
  arrange(desc(total_hits)) %>%
  slice_head(n = 25)

# Extract top motif IDs
top_motifs_ids <- top_motifs$motif_id

# Extract motif names from the JASPAR PWM
motif_names <- sapply(jaspar_pwms, function(x) x@name)
names(motif_names) <- names(jaspar_pwms)

# Filter motif summary to top motif hits
motif_summary_top <- motif_summary %>%
  filter(!is.na(motif_id), motif_id %in% top_motifs_ids) %>%
  mutate(motif_name = motif_names[motif_id])

# Compute the number of TE copies in each TE family
te_family_sizes <- sapply(te_sequences, length)

# Compute the percentage of TE sequences containing each motif
motif_percent <- motif_summary_top %>%
  rowwise() %>%
  mutate(percent_with_motif = n_TE_copies / te_family_sizes[[TE_family]] * 100) %>%
  ungroup() %>%
  select(TE_family, motif_name, percent_with_motif)

# Pivot from long to wide format
motif_matrix_percent <- motif_percent %>%
  pivot_wider(names_from = motif_name, 
              values_from = percent_with_motif, 
              values_fill = 0)

# Convert to numeric matrix
heatmap_mat_percent <- motif_matrix_percent %>%
  column_to_rownames("TE_family") %>%
  as.matrix()

# Plot clustering heatmap showing TF motif abundance in each enriched TE family
colors <- colorRampPalette(c("white", "black"))(100)
pheatmap(
  heatmap_mat_percent,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = FALSE,
  color = colors,
  legend = TRUE,
  fontsize_row = 10,
  fontsize_col = 10,
  main = "Top 25 TF Motifs (% of TE sequences)\n in Enriched TE Families"
)
```
